---
title: "EPA-EIA Crosswalk"
freeze: true
format: html
html:
    toc: true
    toc-expand: true
    toc-location: left
    html-table-processing: none
    code-fold: true
execute: 
  message: false
  warning: false
  error: false
params:
  crosswalk_year: "2018"
  earliest_retirement_year: "2010"
  include_FRS: false
  include_NEEDS: false
  output_agg: "plant"
  unmatch_only: false
project: 
  execute_dir: project
editor: visual
---

# Overview

This project includes all necessary scripts and documentation to create the EPA-EIA crosswalk.

# Background

This R script retrieves the Environmental Protection Agency's (EPA) [Power Sector Emissions Data](https://www.epa.gov/airmarkets/power-sector-emissions-data) and EIA's [860 power plant data](https://www.eia.gov/electricity/data/eia860/) about power plant units and maps the EIA data to EPA data. The EPA data are retrieved with the CAMD [FACT API](https://www.epa.gov/airmarkets/field-audit-checklist-tool-fact-api) and the EIA data are retrieved from the Excel spreadsheets within the ZIP file. You have the option of specifying a different year for EIA data (e.g., the most current year).

The larger data sets from EPA and EIA include information about power plant attributes, emissions, operations, fuel consumption, and electricity generation. This crosswalk relates key identifiers assigned to power plant units to make it easier to connect and use larger EPA and EIA data sets. The crosswalk can also be extended to other government data sets.

Each power plant has a unique facility ID known as an ORIS code (EPA) or plant code (EIA). Steam power plants are made up of boilers (where fuel is combusted to generate steam that moves a turbine) and generators (where kinetic energy from the turbine is converted to electrical energy) to generate electricity. Natural gas-fired power plants are made up of or combustion turbines (where fuel is combusted to turn a turbine) and generators to generate electricity. Combined-cycle power plants also have heat recovery steam generators (where waste heat generates steam that moves a turbine). Boilers, combustion turbines, and generators are interchangeably referred to as units in the two data sets. Because CAMD data is collected for environmental compliance, CAMD considers units to be the combustion units (boilers or combustion turbines). EIA data is collected to track electricity generation, so EIA considers units to be generators. The combustion unit and generator IDs may not be the same, even if they are a part of the same system. In addition, reporting discrepancies may mean that the IDs may not match across the two agencies' databases. This crosswalk attempts to connect these components from both data sets.

# Methodology

## Modify mismatched plant codes

There are some power plants that have a single `plant_id`/`ORIS code` in EPA's data set, but a different or multiple plant codes in EIA's data set. The known manual `plant_id` matches are pulled in from the `manual_matches.xlsx` file. The `eia_plant_id` identifier for these known mismatches are modified and the result is placed in the MOD_EIA_PLANT_ID field (e.g. CAMD PLANT_ID of 330 should match with `eia_plant_id` 57901, so the `mod_eia_plant_id` is changed to 57901 to include the match in the crosswalk). Units are matched using the identifiers preceded by `mod_`

> Note: <br>Steps 1 and 2 are performed independently.
>
> <br>Each sub-step(a-d) within Step 1 and 2 is performed sequentially matching on only the unmatched from the previous sub-step(a-d)
>
> <br>To start, each identifier is copied to the field of the same name preceded by `mod_` and each modification to the original identifier is placed in the corresponding `mod_` field. This means that if a match occurs without modifying the identifier, the `mod_` field will be an exact copy of the original identifier.
>
> <br>Joins are performed on the identifiers preceded by `mod_`

## Step 1: Match EPA to EIA 3_1_Generator data set (retired and operational generators)

Units are matched on ORIS/modified plant code and generator ID.

-   Manual matches and exclusions are pulled in from `manual_matches.xlsx`
-   Step 1a: Units are matched on original unit identifiers
-   Step 1b: Unit identifiers are modified to change to upper case and remove whitespace and special characters.
-   Step 1c: Unit identifiers are modified to change numbers as strings to numeric values.
-   Step 1d: Unit identifiers are modified to remove leading characters if the identifier ends in a number or a number-letter combination (e.g., CT1 becomes 1, CT1A becomes 1A).

## Step 2: Match EPA to EIA 6_1_EnviroAssoc data set (boilers and generators)

Units are matched on ORIS/modified plant code, unit/boiler ID, and generator ID.

-   Manual matches and exclusions are pulled in from `manual_matches.xlsx`
-   Step 2a: Units are matched on original unit identifiers
-   Step 2b: Unit identifiers are modified to change to upper case and remove whitespace and special characters.
-   Step 2c: Unit identifiers are modified to change numbers as strings to numeric values.
-   Step 2d: Unit identifiers are modified to remove leading characters if the identifier ends in a number or a number-letter combination (e.g., CT1 becomes 1, CT1A becomes 1A).

## Step 3: Join data sets from Step 1 and Step 2 to output a set of comprehensive matches that have all EPA identifiers and all EIA identifiers where they exist. EPA units that did not match in any step are added to the crosswalk with an indicator that they were unmatched.

The two sets are joined on `epa_plant_id`, `epa_unit_id`, `epa_generator_id`, `eia_plant_id`, and `eia_generator_id`

> Note: <br>To use this code, you will need an API key. You can request one from the [CAPD API overview](https://www.epa.gov/power-sector/cam-api-portal#/documentation).

# Processing

## Load required packages

```{r}
#| label: load-packages
#| echo: TRUE

library(tidyverse)
# tidyr - Help create tidy data
# stringr - Make working with strings easier (string_glue())
# readr - Read csv, txt, etc
# tibble - Wrapper for data.frames that makes things easier (e.g. never converts strings to factors)

library(lubridate) # Make working with dates easier

library(httr) # Perform HTTP requests (in this case used to get data from FACT API)

library(tidyjson) # Work with json objects in a tidy way. Useful for highly nested objects and "ragged" arrays and/or objects (varying lengths by document)

library(jsonlite)

library(readxl) # Read data from xlsx files via read_excel()

library(openxlsx) # Create and write to formatted xlsx documents

library(purrr) # Use of partial and map functions

```

```{r}
#| label: load-Qmd-packages
#| echo: false

library(docstring)
library(dplyr)
library(ggiraph)
library(ggplot2)
library(gt)
library(gtExtras)
library(kableExtra)
library(knitr)
library(patchwork)
library(readxl)
library(stringr)
```

## Define crosswalk year

The year of the crosswalk specified as the parameter "crosswalk_year" within the YAML of the Quarto document. When a year value needs to be specified (e.g., when pulling relevant data from the EPA API), it is done so by calling `params$crosswalk_year`.

**Current year setting: `r params$crosswalk_year`**

## Load all helper functions

There are a set of helper functions used throughout the project. These are defined in the folder `scripts/functions`. The `docstring` package is used to provide documentation for functions, similar to typical package documentation. To view the documentation for a given function, run `docstring({function_name})`

```{r}
#| label: load-functions
#| echo: false

# sourcing each file in functions folder
functions <- list.files("scripts/functions")
purrr::walk(paste0("scripts/functions/", functions), ~ source(.x))
```

::: panel-tabset
### Overview

*Click on the tab to view the helper function names.*

### Helper Functions

```{r}
#| label: list-functions
#| echo: false

# listing function names
kable(stringr::str_remove(stringr::str_remove(functions, ".R"), "function_"),
      col.names = c("Helper Functions")) %>%
            kable_styling(full_width = F)
```
:::

# Load Data

## Load data from raw sources

### Import EPA data

This section imports the unit and generator data from the EPA CAPD API. **An EPA CAPD API key is required** and can be obtained easily by signing up at the [CAPD API website](https://www.epa.gov/power-sector/cam-api-portal#/documentation). Once a key is obtained, paste your key into the file api_keys/epa_api_key.txt and save. Once saved, the script data_load_epa.R will be able to successfully connect to and load data from the CAPD API.

```{r}
#| label: data-load-epa 
#| file: "scripts/data_load_epa.R" 
#| results: hold
#| echo: false
```

### Import EIA data

This section downloads and imports data from EIA-860 for the year specified above. To manually download the data from EIA, visit the [EIA-860 data](https://www.eia.gov/electricity/data/eia860/). Select and download the latest year's ZIP file on the right-hand-side of the page. The files used in this analysis are **3_1_Generator_Y`r params$crosswalk_year`.xlsx** and **6_1_EnviroAssoc_Y`r params$crosswalk_year`.xlsx**, with **2\_\_\_Plant_Y`r params$crosswalk_year`.xlsx** to get lat/long.

```{r}
#| label: data-load-eia 
#| file: "scripts/data_load_eia.R" 
#| results: hold
#| echo: false
```

# Modify Data

## Modify EIA plant code based on eGRID known mismatch list

In the production of eGRID, CAMD checks whether there are any discrepancies between the ORIS code in CAMD's database and the plant code in EIA's database. A small number of plants do have these mismatches, and they are corrected in this section. This section changes the plant code in EIA's data, placing the changes in the `mod_eia_plant_id` column, to match CAMD's ORIS code data, according to Table C-5 of Appendix C in the [eGRID2018 TSD](https://www.epa.gov/egrid/egrid-technical-support-document). A direct download for this table in xlsx format can be found here: [epa-eia_plant_id_crosswalk.xlsx](https://www.epa.gov/sites/production/files/2020-09/epa-eia_plant_id_crosswalk.xlsx). The manual match Excel file has a copy of the eGRID epa-eia_plant_id_crosswalk that updates. Any updates to the eGRID file will be reflected in the manual match file.

```{r}
#| label: modify-crosswalk-data 
#| file: "scripts/modify_crosswalk.R" 
#| results: hold
#| echo: false
#| eval: false # hold off on running this chunk until later
```

# Match Data

## Match Sub-steps

These steps are performed in sequential order during Steps 1 and 2, the matching process between CAMD and each of the EIA 3_1_Generator (generator) and 6_1_EnviroAssoc (boiler) files.

### Sub-step a - Exact match on all unit identifiers

In this step, we look for direct matches on the EPA and EIA unit identifiers.

### Sub-step b - Modify identifiers to change to uppercase and remove whitespace and special characters

In this step, we modify the EPA and EIA unit identifiers to adjust all alpha characters to uppercase, remove whitespace, and remove special characters Examples of these modifications include:

-   un14 -\> UN14
-   CT 2A -\> CT2A
-   CT-1 -\> CT1

### Sub-step c - Modify identifiers to convert numbers in strings to numeric values

In this step, we modify the EPA and EIA unit identifiers to convert all numbers presented as strings to numeric values. This also eliminates leading zeros. Examples of these modifications include:

-   002 -\> 2
-   "2" -\> 2

### Sub-step d - Modify identifiers to eliminate leading alpha characters

In this step, we modify the EPA and EIA unit identifiers to remove all alpha (letter) characters at the start of the identifier if the identifier ends in a number or number-letter combination. Examples of these modifications include:

-   CT1A -\> 1A
-   SGT1 -\> 1
-   No 5 -\> 5

## Steps 1, 2 and 3:

Look for matches on the modified EPA and EIA unit identifiers between CAMD and EIA 3_1_Generator (generators) EIA 6_1_EnviroAssoc (boilers) separately. After both matches are complete between CAMD and each of the EIA data sets, we join the matches to have a result with matches between:

`epa_plant_id` = `eia_plant_id`

`epa_unit_id` = `eia_boiler_id`

`epa_generator_id` = `eia_generator_id`

```{r}
#| label: match-crosswalk-data 
#| file: "scripts/match_crosswalk.R" 
#| echo: false
```

```{r}
#| label: generator-match-table
#| echo: false

generator_match_table <- generator_match_summary %>%
                         gt::gt() %>%
                         gt::tab_style(
                              style = cell_text(size = 14, weight = "bold"),
                              locations = cells_column_labels()) %>%
                         gt::tab_caption(caption = "Generator Match Summary")

generator_match_table
```

```{r}
#| label: boiler-match-table
#| echo: false

boiler_match_table <- boiler_match_summary %>%
                      gt::gt() %>%
                      gt::tab_style(
                            style = cell_text(size = 14, weight = "bold"),
                            locations = cells_column_labels()) %>%
                      gt::tab_caption(caption = "Boiler Match Summary")


boiler_match_table
```


### Match Summary

`epa_distinct`/`epa_duplicates` are found according to unique combinations of EPA plant, unit, and generator identifiers. Duplicates can either be introduced from the automated crosswalk operations or from manual matches. Duplicates can also occur in the rare cases where CAMD has fewer identifiers for the same unit/generator than EIA (e.g. `epa_plant_id` = 52151, `epa_unit_id` = 001, `eia_plant_id` = 52151, `eia_boiler_id` = PB1,RF1,PB2,RF2).

```{r}
#| echo: false
#| label: crosswalk-summary-table

# Create table of counts of matches for each step
crosswalk_summary_table <-
  epa_eia_crosswalk %>%
  filter(str_detect(match_type_gen, "Unmatched|Excluded", negate = T) | str_detect(match_type_boiler, "Unmatched|Excluded", negate = T)) %>%
  summarize(
    epa_matched = n(),
    epa_distinct = n_distinct(epa_plant_id, epa_unit_id, epa_generator_id),
    epa_duplicate = n() - epa_distinct
  ) %>%
  mutate(
    epa_unmatched = epa_eia_crosswalk %>%
      filter(str_detect(match_type_gen, "Unmatched", negate = F) | str_detect(match_type_boiler, "Unmatched", negate = F)) %>%
      nrow(),
    epa_excluded = epa_eia_crosswalk %>%
      filter(str_detect(match_type_gen, "Excluded", negate = F) | str_detect(match_type_boiler, "Excluded", negate = F)) %>%
      nrow()
  ) %>%
  gt::gt() %>%
  gt::tab_style(
      style = cell_text(size = 14, weight = "bold"),
      locations = cells_column_labels()) %>%
  gt::tab_caption(caption = "Crosswalk Summary")

crosswalk_summary_table
```

## Add Optional Datasets

### FRS Data

```{r}
#| label: import-frs-data
#| results: hold
#| echo: TRUE
# add FRS IDs to crosswalk output
# set to run/not run by changing TRUE/FALSE at the beginning of the code
# WARNING: could take several minutes to run due to data retrieval

if(!params$include_FRS) {
  print("No FRS IDs added")
} else {
  source("scripts/import_frs.R")
  print("FRS IDs added")
}
```

### NEEDS Data

```{r}
#| label: import-needs-data
#| results: hold
#| echo: TRUE

if(!params$include_NEEDS) {
  print("No NEEDS IDs added")
} else {
  source("scripts/import_needs.R")
  print("NEEDS IDs added")
}
```

## Crosswalk Output

This document outputs an Excel file and a CSV file aggregated based on the params input `output_agg`.

**Current output aggregation: `r params$output_agg`**

```{r}
#| label: output-crosswalk
#| file: "scripts/output_crosswalk.R" 
#| results: hold
#| echo: false
```
